Revealing the rich and varied dynamics of hippocampal replay

### 2020-07-29 Article.tex
Draft 2

\title{A state space model reveals the rich and varied dynamics of hippocampal replay}
\shorttitle{Replay Dynamics}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[5]{Marielena Sosa}
\author[6]{Kenneth Kay}
\author[7]{Jason E. Chung}
\author[4]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

\affil[1]{Howard Hughes Medical Institute, University of California, San Francisco, San Francisco, California}
\affil[2]{Department of Physiology, University of California, San Francisco, San Francisco, California}
\affil[3]{Kavli Institute for Fundamental Neuroscience, University of California, San Francisco, San Francisco, California}
\affil[4]{Department of Mathematics and Statistics, Boston University, Boston, Massachusetts}
\affil[5]{Department of Neurobiology, Stanford University, Palo Alto, California}
\affil[6]{Mortimer B. Zuckerman Mind Brain Behavior Institute, Columbia University, New York, New York}
\affil[7]{Department of Neurological Surgery, University of California, San Francisco, California}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past or possible future experiences play a critical role in memory and decision-making processes. The hippocampus is a brain region known to express these representations across multiple states, including during sharp-wave ripple (SWR) events where sets of hippocampal place cells can be activated in time-compressed sequences consistent with a rapid mental traversal of a path. Efforts to understand these "replay" events typically analyze a small subset of events and assume rapid mental movement, and as a result we lack a clear picture of both the prevalence and the range of dynamics associated with replay. Here we develop a flexible state space model that uses a combination of discrete and latent states to capture the content and time evolution of SWR events. Application of this model reveals the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

\section*{Introduction}
The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These internal "generative" representations \cite{KayConstantSubsecondCycling2020} can be identified because the underlying neural activity patterns resemble those seen during actual events. In particular, an encoding model that relates neural firing to external covariates can be constructed during an experience, and this model can be inverted to decode an internal representation of the covariates from the neural activity at a different time.

@@

Hippocampal replay events are a prototypical example of a generative representation. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells can fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience or likely to occur during potential future experiences, and previous work has reported that these sequential firing events proceed at an average speed of ~10 meters / second, about 20x faster than the animal's usual movement speed \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

Identifying these events typically involves multiple steps and assumptions about their nature. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which instantiates the assumption that the representations progress at a constant speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each even \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which both approaches often exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} and characterizes represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 rats, focusing on SWR events to permit a direct comparison to previous work.

We find that the vast majority of SWRs contain spatially coherent content. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at speeds consistent with real animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} 19 simulated place cells that exhibit three different sequence patterns. For the first third of time, one cell fires repeatedly, representing one stationary location. For the second third of time, the cells fire in sequence, representing a smooth continuous trajectory through space. For the last third cells fire in an incoherent pattern, representing a non-smooth fragmented trajectory through space. \textbf{(B)} Like the Bayesian Decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} This is combined with an explicit model of each dynamic state--which determines how latent position can change based on the posterior from the previous time step. In the stationary state, this is modeled as the identity matrix because the position remains the same. In the continuous state, the dynamics are modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close. In the fragmented state, the dynamics are modeled as uniform, meaning the next position is equally likely to be anywhere. Note that the dynamics displayed only reflect the within-state dynamics. \textbf{(D)} The probability of remaining in a particular state versus switching to another state. This is modeled as having a high probability of remaining in a particular state with a small probability of switching to one of the other dynamics at each time step. \textbf{(F)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time (lines). This can also be used to get an estimate of latent position over time (right panel). \textbf{(G)} The probability of each state given a particular constant speed trajectory. Each speed corresponds to a simulated spiking sequence at that speed. The lines correspond to the average probability of that state over the sequence. Dotted line represents the 0.8 classification threshold we use to classify each range of speeds and shaded regions correspond to the classification we have given the range of speeds.
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then apply our decoding algorithm to the spiking sequence in Figure \ref{1}A. For the first third of this sequence, a single place cell fires repeatedly, representing a single location. For the middle third of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory. For the last third, the cells fire in a incoherent spatial order. These firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamics model---how the position can change over time. For the data model, our decoder is the same as the Bayesian decoder. We compute an estimate of how each cell's firing rate varies over position (the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the Bayesian decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the Bayesian decoder.

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells---evolves over time (Figure \ref{1}C). We do this by defining a movement state transition matrix, which defines how the latent position can change from the previous time step. Previous findings suggest three distinct dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, smoothly moving \cite{DavidsonHippocampalReplayExtended2009}, and fragmented / jumpy \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both spatially incoherent representations and representations that jump from one position to another. We therefore developed state transition models to capture each of these dynamics.

In the stationary dynamic, the latent position would not change between time steps, so we use an identity state transition matrix, which predicts the next position will be the same as the last position. In the continuous dynamic, the latent position would be "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since we are interested in identifying replays that move at speeds much faster than the animal's speed, we set this to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (corresponding to a latent speed of up to ~25 m/s). This is consistent with replay speeds observed in previous studies \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}. Finally, in the fragmented dynamic, the latent position could move to any available position instantaneously. We model this using a uniform state transition matrix, which makes all positions equally likely.

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for \texttildelow100 ms, which is approximately the duration of the SWR. There is, however, a small probability of switching to one of the other dynamics (note that below we show that our results are relatively insensitive to this parameter). Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drives the estimated dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic.

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic state (see Methods). With this, we can summarize the resulting posterior with two quantities: the mixture probability of each dynamic over time and the probability of position over time, irrespective of dynamic (Figure \ref{1}F, left and right plot respectively).

An examination of these two summaries reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and the probability of the stationary dynamic is high at the beginning of the simulated replay. The change in the data then drives a rapid transition to the continuous dynamic, reflecting the trajectory-like spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented dynamic dominates for the last third. Importantly, these dynamics also inform the estimate of latent position of the animal, and the correct assignment of dynamic reduces the variance of the latent position estimate.

The specification of these dynamics allows the model to capture a wide range of movement speeds for the latent position. To demonstrate this, we applied the model to simulated trajectories of increasing speed (Figure \ref{1}G), from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three dynamics being high probable (where we define highly probable to be over 0.8 probability), there are intermediate speeds where two of the dynamics dominate with high probability; that is, the sum of two of the dynamics' probabilities is over 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary state has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its dynamics.

We note here that the choice of any particular threshold is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95 because the 0.95 threshold often leads people to conclude that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v4}
\caption{Demonstration of the state space model on real data using both sorted and clusterless spikes on the same SWR. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure 1F. Shaded regions correspond to the speed classifications as in Figure 1G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as(A).  \textbf{(C)} Maximum of the 1D decoded position (the most probable position) in (A) projected back into 2D. Color indicates time. The animal's actual position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire epoch. L, R, and C correspond to the left, right and center reward well as in (A). \textbf{(D)} Maximum of the 1D decoded position (the most probable position) in (B) projected back into 2D. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v4}
\caption{A-D. Four examples of replays that would not be well-characterized by using the Bayesian decoder. Figure conventions are the same as in Figure 2. \textbf{(A)} A replay that starts down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that stays at the left well for the entirety of the SWR while the animal is in the center arm. \textbf{(C)} A replay that begins stationary at the left well and then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins stationary at the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories (Stationary, Stationary-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Stationary, Stationary-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{3}
\end{figure*}

>>>>

\title{Revealing the rich and varied dynamics of hippocampal replay}
\shorttitle{Replay Dynamics}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[6]{Kenneth Kay}
\author[5]{Marielena Sosa}
\author[7]{Jason E. Chung}
\author[4]{Uri T. Eden}
\author[1, 2, 3, \Letter]{Loren M. Frank}

\affil[1]{Howard Hughes Medical Institute, University of California, San Francisco, San Francisco, California}
\affil[2]{Department of Physiology, University of California, San Francisco, San Francisco, California}
\affil[3]{Kavli Institute for Fundamental Neuroscience, University of California, San Francisco, San Francisco, California}
\affil[4]{Department of Mathematics and Statistics, Boston University, Boston, Massachusetts}
\affil[5]{Department of Neurobiology, Stanford University School of Medicine, Stanford, CA}
\affil[6]{Mortimer B. Zuckerman Mind Brain Behavior Institute, Columbia University, New York, New York}
\affil[7]{Department of Neurological Surgery, University of California, San Francisco, California}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on event significance, and thus analyze only a small subset of events.  As a result, we lack a clear picture of both the prevalence and the range of dynamics associated with replay. Here, we develop a flexible state space model that uses a combination of discrete and latent states to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that the large majority of SWR events contain coherent spatial content that evolves in accordance with a surprisingly wide range of mental movement speeds.

@@

\section*{Introduction}

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These internal "generative" representations \cite{KayConstantSubsecondCycling2020} can be identified because the underlying neural activity patterns resemble those seen during actual events. In particular, an encoding model that relates neural firing to external covariates can be constructed during an experience, and this model can be inverted to decode an internal representation of the covariates from the neural activity at a different time.

@@

Hippocampal replay events are a prototypical example of a generative representation. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience or likely to occur during potential future experiences, and previous work has reported that these sequential firing events proceed at an average speed of ~10 meters / second, about 20x faster than the animal's usual movement speed \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

Identifying these events typically involves multiple steps and assumptions about the nature of replay, which are mostly commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations smoothly progress at a constant speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using some arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, which both approaches often exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 rats, focusing on SWR events to permit a direct comparison to previous work.

We find that the majority of SWRs contain spatially coherent content; that is, the replay trajectories have no large jumps. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with real world animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

\section*{Results}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure1/Figure1_v4}
\caption{Components of the state space model and demonstration on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For the second 60 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 60 ms, cells fire randomly, out of their spatial order, representing an incoherent spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the posterior from the previous time step. In the stationary movement dynamic, the transition matrix is modeled as the identity matrix because the position remains the same. In the continuous movement dynamic, the transition matrix is modeled as a Gaussian random walk, meaning the next possible position is likely to be spatially close to the previous position. In the fragmented movement dynamic, the transition matrix is modeled as uniform, meaning the next position is equally likely to be any possible position. Note that the movement dynamics displayed only reflect the within-state movement dynamics. \textbf{(D)} The probability of remaining in a particular dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The state space model uses the components in A-D over all time to decode the joint probability of latent position and dynamic state. The joint probability can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic state over time. The summed posterior across states also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. The lines correspond to the average probability of that dynamic for a given constant velocity sequence. We use a 0.8  threshold (dotted line) to classify each sequence based on the state or states which contribute maximally to the posterior. 
}
\label{1}
\end{figure*}
\subsection*{Overview of the model}
We begin with an application of the model to simulated data, both to validate the model and to provide intuition (Figure \ref{1}). We simulated 19 Poisson spiking cells with Gaussian place fields on a 180 cm virtual linear track. Each place field has a 36 cm variance and a 15 Hz peak firing rate, which is spaced every 10 cm along the virtual track. We then construct an 180 ms spiking sequence and apply our decoding algorithm in Figure \ref{1}A. For the first 60 ms of this sequence, a single place cell fires repeatedly, resulting in the extended representation of a single location. For the second 60 ms of the sequence, the cells fire in sequential spatial order, representing a fast moving trajectory across the virtual linear track. For the last 60 ms of the sequence, the cells fire in an incoherent spatial order. These three firing patterns represent three different types of sequence dynamics that could be expressed during replay events, which we call stationary, continuous, and fragmented, respectively. The goal of our model is to characterize SWRs in terms of a mixture of these three dynamics at every time point.

Decoding the spiking sequence dynamics requires specifying two elements: the data model---how the spikes relate to position---and the movement dynamics model---how the position can change over time. For the data model, our decoder is the same as the standard ("Bayesian") decoder \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015, StellaHippocampalReactivationRandom2019}. We compute an estimate of how each cell's firing rate varies over position during movement (i.e. the place field, Figure \ref{1}B). This is used during decoding to compute the Poisson likelihood of position over time given the spiking sequence of interest. In contrast to the standard decoder, we can use arbitrarily small time bins (in our case 2 ms vs. 20 ms or more) because we are able to take advantage of the prior placed on the dynamics by the state space model. This allows us to detect changes on smaller time scales than would be possible with the standard decoder.

Next, we specify a movement dynamics model for how latent position---the "mental" position of the animal represented by the cells---evolves over time (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, rapid continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented \cite{PfeifferAutoassociativedynamicsgeneration2015}, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another. We therefore define state transition models to capture each of these dynamics.

In the stationary dynamic, the latent position does not change between time steps. The state transition matrix can thus be defined as an identity matrix, which predicts the next position will be the same as the last position. In the continuous dynamic, the latent position is most likely to be  "spatially close" to the position in the previous time step, so we use a Gaussian random walk state transition matrix. This means that, for a given latent position, the probability of moving to another position is modeled by a Gaussian centered at that position and "spatially close" is defined by the variance of the Gaussian. In our case, since replay has been shown to move at speeds much faster than the animal's speed \cite{DavidsonHippocampalReplayExtended2009, PfeifferAutoassociativedynamicsgeneration2015}, we set the variance of the Gaussian to 6.0 cm. This ensures that with a 2 ms time step, the latent position is 95\% likely to be within 4.90 cm of the previous latent position (or equivalently, this means that latent speeds from 0 to ~25 m/s are most likely).  Finally, in the fragmented dynamic, the latent position can move to any available position instantaneously. We model this using a uniform state transition matrix, which makes transitions to all positions equally likely.

Finally, we specify how likely each dynamic is to persist in time versus changing to another dynamic (Figure \ref{1}D). In order to be conservative, we assume that each dynamic is likely to dominate for \texttildelow100 ms, which is approximately the duration of a SWR event. There is, however, a small probability of switching to one of the other dynamics. Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step. Because the probability of changing dynamics follows a geometric distribution, this corresponds to an expected duration of 100 ms for staying in a particular dynamic. Importantly, the data drives the estimated dynamic, so even if the probability of a particular dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic. In line with this, note that below we show that our results are relatively insensitive to the value of this parameter.

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior with two quantities: the mixture probability of each dynamic over time (by marginalizing over position) and the probability of latent position over time, irrespective of dynamic (by marginalizing over dynamic; Figure \ref{1}F, left and right plot respectively).

An examination of the two summaries shown in Figure \ref{1}F reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and accordingly, the probability of the stationary dynamic is high at the beginning of the simulated replay. A change in the data then drives a rapid transition to the continuous dynamic, reflecting the trajectory-like spatially sequential spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented dynamic dominates for the last third of the simulated event. Importantly, these dynamics also inform the estimate of latent position of the animal, and the correct assignment of dynamic reduces the variance of the latent position estimate.

The specification of these dynamics allows the model to capture a wide range of movement speeds for the latent position. To demonstrate this, we applied the model to simulated trajectories of increasing speed (Figure \ref{1}G), from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three dynamics being highly probable (where we define highly probable to be over 0.8 probability), there are intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary state has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G). By using this classification scheme, we can characterize the speed or set of speeds in a replay by its dynamics.

We note here that the choice of any particular threshold for classifying a SWR is arbitrary, and that the power of our approach lies in part in the ability to assign a confidence value for each dynamic or combinations of dynamics to each moment in time. Our goal in choosing 0.8 was to use a threshold that corresponds to highly probable and to explicitly avoid the arbitrary choice of 0.95, as the 0.95 threshold risks giving the false conclusion that only events that are statistically significant at the 0.05 level are meaningful in the brain. Nonetheless, we also verified that our central results hold with a higher threshold of 0.95.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v4}
\caption{Demonstration of the state space model on hippocampal data using both sorted and clusterless spikes on the same SWR event. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure \ref{1}F. Shaded regions correspond to the speed classifications as in Figure \ref{1}G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the red triangle. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as (A).  \textbf{(C)} Maximum of the 1D decoded position (the most probable position) in (A) projected back onto the 2D maze for the sorted decoding. Color indicates time. The animal's actual position is denoted by the red circle. Light grey lines show the animal's 2D position over the entire recording session. L, R, and C correspond to the left, right and center reward well as in (A). \textbf{(D)} Maximum of the 1D decoded position (the most probable position) in (B) projected back into 2D using the clusterless decoding. Figure conventions are the same as in (C).
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v4}
\caption{A-D. Four examples of replays that would not be well-characterized by using the standard decoder. Figure conventions are the same as in Figure \ref{2}. \textbf{(A)} A replay that starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A replay that persistently represents the left well for the entirety of the SWR while the animal remains in the center arm. \textbf{(C)} A replay that begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A replay that begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} Percentage of SWRs in each epoch that has some content that falls into one of the categories of dynamics (Stationary, Stationary-Continuous-Mix, Continuous, Fragmented-Continuous-Mix, or Fragmented). \textbf{(F)} Of the classified content, the percentage of SWRs in each epoch that have some Stationary, Stationary-Continuous-Mix, or Continuous content. \textbf{(G)} Of the classified content, the percentage of SWR in each epoch that has some Continuous content.
}
\label{3}
\end{figure*}

We next sought to validate the model on real hippocampal data. We fit the place fields of 31 cells recorded in hippocampus while the rat was performing a spatial alternation task on a W-shaped track, and then applied the decoding algorithm to a SWR with sequential population activity (Figure \ref{2}A, top panel). As expected, we observed that the probability of being in the continuous dynamic is high throughout this event, but the probability of being in a stationary state was also noticeable at the beginning and end of the SWR (Figure \ref{2}A, middle panel). Using our speed classification scheme, this means that the speed of the replay starts slower---as a mixture of continuous and stationary dynamics---and then speeds up and slows down again. This is also evident in the posterior probability of linear position over time. This shows that the replay most likely travels down the center arm and up the right arm (Figure \ref{2}A, bottom panel). We can also see this when we project the maximum of the posterior of this trajectory (the most probable "mental" position) to 2D (Figure \ref{2}C) to better see the spatial trajectory on the maze. Importantly, when we apply the same model using the 2D position of the animal, we get a similar result as when we use the 1D linearized position (Figure \ref{fig:Figure2-Figure supplement 1}A).
